NumPy---Array Computation for Python

### 2019-12-19 paper.tex
Add notes
Rework history, funding, and current status sections (w/ Jarrod Millman)

@@

NumPy was, at its inception, the unifier of the Python array
processing world.  Before NumPy, two packages—Numeric and
numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
small arrays.  In contrast, to handle large astronomical images,
NumArray was developed at the Space Telescope Science Institute (STScI).
In other words, these array
objects had different strengths and weaknesses, and therefore tended
to divide the already small user community.  In 2005, Travis Oliphant
started NumPy, a ``best of both worlds'' unification of the two
projects, based on Numeric.  The arrival of stable version 1.0 in October
2006 unified the scientific Python
community, and heralded the era of popular scientific computing in
Python that has led to formidable successes such as [cite some impacts here].

While, by 2006, NumPy was already eminently usable, it lacked good
documentation. In order to use NumPy in his classes, Joe Harrington
from the University of South Florida sponsored several rounds of the
NumPy Documentation Marathon \cite{harrington2008scipy,vanderwalt2008scipy}.
During these intensive
documentation improvement sessions, a Wikipedia-like system was
developed to allow community members to contribute so-called
``docstrings''—text describing each NumPy function. A community
documentation standard was published so that these pieces of
documentation would all provide similar information, and conform to a
basic standard of quality. Experienced community members reviewed and
integrated contributed docstrings, and soon NumPy had thorough
documentation coverage.

For a more detailed history of the scientific Python ecosystem, please
refer to ``SciPy 1.0---Fundamental Algorithms for Scientific Computing
in Python'' \cite{virtanen2019scipy}.

@@

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
and adoption without external investment.

@@

The Berkeley Institute for Data Science (BIDS) received two grants to support
the development of NumPy.
At the end of 2016, the Gordon and Betty Moore Foundation awarded them
\$645,020 for improving NumPy for better data science.
In 2017, the Alfred P. Sloan Foundation complemented the funding with an
additional \$659,359.
Stéfan van der Walt, a senior research data scientist at BIDS and a member of
the NumPy steering council, is the PI of both grants and manages a team of
four programmers working on the project.

@@

NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
and OpenBLAS from the Chan Zuckerberg Initiative. 
NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
Ralf Gommers, the Director of Quansight Labs and a member of
the NumPy steering council, is the PI.

% focus on better serve NumPy's large number of beginner to intermediate level users
% focusing all our time on growing the team and better serving new users and contributors

@@

Over the past few years, with external financial support, NumPy has seen
renewed developer interest.
It is actively developed, with weekly online developer
meetings and a formalized NumPy enhancement proposal process.

Following Python's practice of using Python Enhancement Proposals (PEPs) for
``proposing major new features, for collecting community input on an
issue, and for documenting the design decisions that have gone into Python,''
we began using NumPy Enhancement Proposals (NEPs) in 2007.
This process was informal and only occasionally used until 2017 when
we adopted a more formal process \footnote{\url{https://numpy.org/neps/nep-0000.html}}.
Since then there have been 19 proposed NEPS---6 have been implemented,
4 have been accepted and are being implemented, 4 are under consideration,
3 have deferred or superseded, and 2 have been rejected or withdrawn.
We highlight a few of the implemented NEPs below.

NumPy plays a central role in building and standardizing much of the scientific
Python community infrastructure.
The docstring standard mentioned in the history section is now widely adopted.
We are also using the NEP system as a way to help coordinate the larger
scientific Python community.
% https://numpy.org/neps/nep-0029-deprecation_policy.html
For example, in NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem adopt a
common ``time window-based'' policy for support of Python and NumPy versions.
This standard will simplify downstream project and release planning.

@@

@@

NumPy provides a large library of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
For historical reasons, NumPy also includes basic functionality for
linear algebra,
fast Fourier transforms and windowing,
pseudorandom number generators,
and polynomial fitting.
We will keep supporting these, but we will not expand beyond them.

@@

%NumPy's linear algebra library includes functions for: solving linear
%systems of equations; computing various functions of a matrix, including
%the determinant, the norm, the inverse, and the pseudo-inverse;
%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
%and more.
%
%The random number generator library in NumPy provides several alternative
%\emph{bit stream generators} that provide the core function of generating
%random integers.
%A higher-level generator class that implements an assortment of
%probability distributions is provided. It includes the beta, gamma
%and Weibull distributions, the univariate and multivariate normal
%distributions, and more.

@@

@@

@@

@@

>>>>

The arrival of a stable version of NumPy (1.0) in October 2006 unified
the scientific Python community, and heralded the era of popular
scientific computing in Python that has led to formidable successes
such as [cite some impacts here].

[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
  IDL is now below Python, Pangeo <-> xarray, Earthcube,
  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

@@

NumPy was, at its inception, the unifier of the Python array
processing world.  Before NumPy, two packages—Numeric and
numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
small arrays.  In contrast, to handle large astronomical images,
NumArray was developed at the Space Telescope Science Institute (STScI).
In other words, these array
objects had different strengths and weaknesses, and therefore tended
to divide the already small user community.  Furthermore, the divide
posed a challenge for SciPy, which built on Numeric and provided
fundamental scientific routines \cite{virtanen2019scipy}.
In response, the maintainer of Numeric, Travis Oliphant, started a
major rewrite that aimed to be a ``best of both worlds'' unification
of the two projects.  This became known as NumPy.

While, by 2006, NumPy was already eminently usable, it lacked good
documentation. In order to use NumPy and SciPy in his classes, Joe Harrington
from the University of South Florida sponsored several rounds of the
NumPy Documentation Marathon \cite{harrington2008scipy,vanderwalt2008scipy}.
During these intensive
documentation improvement sessions, we developed a Wikipedia-like
system for community members to contribute so-called
``docstrings''—text describing each NumPy function. A community
documentation standard was published so that these pieces of
documentation would all provide similar information, and conform to a
basic standard of quality. Experienced community members reviewed and
integrated contributed docstrings, and soon NumPy had thorough
documentation coverage.

% developed best practices, version control, unit testing, code review,
% runable documentation, \cite{millman2014developing}
% Travis steps back, no focused community effort until grant

To coordinate work as a global community, to ensure accuracy of
computed results, and to allow modification without breakage, the
NumPy community pioneered best software engineering practices in the
scientific Python community.  For example, NumPy early on adopted
distributed revision control, unit testing, continuous integration,
code review, standardized executable documentation, ...

However, after his initial effort to build NumPy, Travis Oliphant
stepped back from development to found a company and several projects,
including Numba---an accelerated Python compiler that could vastly
accelerate NumPy expressions.  By 2009, the project had entered a
period of uncoordinated, opportunistic development, and was
increasingly weighed down by technical debt that had accrued since the
days of Numeric.

@@

% built up technical debt, from a legacy of code from two projects
% and a lack of fulltime attention left ...

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the
package had rapid adoption and became central to scientific computing
without external investment.

% From grant text/summary on Sloan website:
%
% to remove the technical debt and set in place standards and
% architecture to encourage more sustainable development

@@

In 2017, the Berkeley Institute for Data Science
(BIDS) received grants totaling 1.3M USD from the Gordon \& Betty
Moore and the Alfred P. Sloan foundations to support the development
of NumPy for better data science.
Specifically, the aim was to address technical debt and set in place
standards and architecture to encourage more sustainable development.
Stéfan van der Walt, a senior research data scientist at BIDS and a
member of the NumPy steering council, is the PI and manages four
programmers working on the project.

@@

In 2019, NumFOCUS received a grant for \$195,000 in 2019 for the
development of NumPy and OpenBLAS (on which NumPy depends for
accelerated linear algebra) from the Chan Zuckerberg Initiative.
This grant, building on the technical work supported by the previous
grant, will focus on better serving NumPy's large number of beginner
to intermediate level users and on growing the community of NumPy
contributors.
Ralf Gommers, the Director of Quansight Labs and a member of the NumPy
steering council, is the PI.

The project receives around 10K USD per year from Tidelift, which is
used to fund documentation and website developers.

@@

In 2007, the NumPy release manager, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently until 2017 when, to
facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
Since then there have been 19 proposed NEPS---6 have been implemented,
4 have been accepted and are being implemented, 4 are under
consideration, 3 have been deferred or superseded, and 2 have been rejected
or withdrawn.

The first activities organized for the new grant were a NEP development
sprint, a weekly community call, and a meeting to write a draft roadmap.

Since the roadmap was meant to represent community consensus on future
development, it was presented for comment at the annual SciPy
conference during a so-called ``Birds of a Feather'' session, attended
by more than 100 people.  One additional meeting was held to discuss
specific enhancements, such as the development of a new data-type
system.  The final roadmap was vetted by the community via online
discussion before being accepted.

The weekly community calls alternate weekly between triage meetings and
higher level discussion.  The calls not only involve developers from
the community, but provide a venue for vendors and other external
groups to provide input.  For example, after Intel produced a forked
version of NumPy, one of their developers joined a call to discuss
community concerns.

NumPy plays a central role in building and standardizing much of the scientific
Python community infrastructure.
The docstring standard mentioned in the history section is now widely adopted.
We are also now using the NEP system as a way to help coordinate the larger
scientific Python community.
% https://numpy.org/neps/nep-0029-deprecation_policy.html
For example, in NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem adopt a
common ``time window-based'' policy for support of Python and NumPy versions.
This standard will simplify downstream project and release planning.

@@

We showed that NumPy is actively developed according to a community
vetted plan that integrates formal enhancement proposals.  Next, we
give a more in-depth overview of the team developing NumPy, their
activities, as well as the governance and community structures that
guide the collaboration.

@@

NumPy provides a large library of functions for array manipulation
including functions for: creating, reshaping, concatenating, and padding arrays;
searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.
For historical reasons, NumPy also includes basic functionality for
linear algebra,
fast Fourier transforms and windowing,
and polynomial fitting.
We will keep supporting these, but we will not expand beyond them.

% use scipy for linear algebra etc...

@@

%NumPy's linear algebra library includes functions for: solving linear
%systems of equations; computing various functions of a matrix, including
%the determinant, the norm, the inverse, and the pseudo-inverse;
%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
%and more.
%

@@

%The random number generator library in NumPy provides several alternative
%\emph{bit stream generators} that provide the core function of generating
%random integers.
%A higher-level generator class that implements an assortment of
%probability distributions is provided. It includes the beta, gamma
%and Weibull distributions, the univariate and multivariate normal
%distributions, and more.

@@

% Mention explicitly the NEPs involved

@@

% started at as focused small matrices, class examples

@@

% 1st grant done, technical cleanup finished except dtypes
% 2nd grant starting, community growth and new user focus
